Avoiding accuracy-limiting pitfalls in the study of protein-ligand interactions with isothermal titration calorimetry

### 2015-07-10 manuscript/itc-worksheet.tex
Updated ITC manuscript with SEB edits. Added NBSIR reference.

\author[qb3]{Sarah E.~Boyce\fnref{gilead}}
%\ead{sarah.boyce@gilead.com}
%\ead[url]{home page}
%\fntext[label2]{label2}
%\cortext[cor1]{cor1}
%\address[qb3]{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720}
\fntext[gilead]{Current address: Gilead Sciences, Foster City, CA 94404}

@@

\author[qb3]{John D.~Chodera\corref{cor1}}
%\fntext[label2]{label2}
\cortext[cor1]{Corresponding author.}
\ead{jchodera@berkeley.edu}
%\ead[url]{http://choderalab.org}

\address[qb3]{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720}
\address[vanderbilt]{Department of Chemistry, Vanderbilt University, Nashville, TN 37235}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NO ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Isothermal titration calorimetry (ITC) can yield precise ($\pm$3\%) estimates of the thermodynamic parameters describing biomolecular association (affinity, enthalpy, and entropy), making it an indispensable tool for biochemistry and drug discovery.  
Surprisingly, interlaboratory comparisons suggest that errors of $\sim$20\% are common and widely underreported. 
Here, we show how to reduce precision- and accuracy-limiting errors while obtaining good estimates while minimizing material and time consumed by an experiment.
We provide a simple spreadsheet that allows practitioners to identify precision-limiting operations during protocol design, track precision during the experiment, and propagate error to yield realistic final uncertainties.\\
\emph{Keywords: isothermal titration calorimetry (ITC), propagation of error, entropy-enthalpy compensation}
\end{abstract}

@@

For illustration, we consider the target reaction from the ABRF-MIRG'02 survey~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}, the 1:1 association of CBS and bovine CAII, which can be written,
\begin{equation}
M + X \stackrel{K}{\leftrightharpoons} MX \label{equation:binding-reaction} ,
\end{equation}
where $M$ denotes macromolecule and $X$ ligand.
This reaction has a $K_a \sim 10^6$M and $\Delta H \sim -10$ kcal/mol~\cite{myszka:2003:j-biomol-tech:abrf-mirg02,tellinghuisen-chodera:2011:anal-biochem:systematic-itc-errors}.

As both protein and ligand may be precious, there is a desire to minimize material use in protein-ligand studies.
Using concentrations only as large as necessary also minimizes the need for buffer additives such as DMSO to enhance solubility, reducing agents to prevent crosslinking, and detergents to inhibit aggregation. 
These additives pose additional experimental challenges, as calorimetrically-measured heats can be sensitive to even small composition mismatches between cell and syringe solutions. 
Minimizing these effects requires dialysis of the macromolecule by buffer followed by preparation of the ligand in the dialysate.  
If the ligand is already in solution (e.g.~in DMSO stock), it may not be possible to fully eliminate excipients, leading to potential heat effects due to buffer mismatch even if attempts are made to match compositions~\cite{tellinghuisen-boyce-chodera:2012:anal-biochem:blanks}.

In the ABRF-MIRG'02 survey~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}, participants employed titrand (cell reagent) concentrations [M]$_0$ in the range 7--71 $\mu$M.
We used an ITC protocol design program~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}, which indicated $\sim$3\% relative squared error (RSE) in $K$ and $\sim$1\% for $\Delta H$ was possible with our instrument (a GE/MicroCal VP-ITC) using [M]$_0$ = 10 $\mu$M (consuming $\sim$0.5 mg protein per experiment).
While this gives $c = K [M]_0 \approx 10$, a key ITC parameter~\cite{wiseman:anal-biochem:1989:itc}, in the low range of that recommended by standard protocols, high measurement precision can still be obtained at this $c$ value by titrating to an optimal titrant:titrand ratio $R_m$ given by,
\begin{equation}
R_m = \frac{6.4}{c^{0.2}} +\frac{13}{c} \label{equation:limiting-concentration},
\end{equation}
a heuristic expression\footnote{While use of this expression requires a rough estimate of the reaction $K$ and an $[M]_0$ that will produce observable heats, this is currently unavoidable in the practice of calorimetry. In the worst case, a pilot experiment using minimal material can be used to crudely estimate these quantities and Eq.~\ref{equation:limiting-concentration} used to determine optimal conditions for a second experiment.} obtained from a comprehensive study of precision as a function of $R_m$~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc}.
The suggested $R_m = 5.3$ is significantly greater than the $R_m = 2$ that is widely used in standard protocols for ITC; with decreasing $c$, use of $R_m = 2$ progressively limits the fractional conversion of M to MX and thus limits the precision of estimation for both $K$ and $\Delta H$~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}.
In the present case, use of $R_m = 2$ would cause significant precision loss, almost doubling the achievable RSEs for $K$ and $\Delta H$.

The same optimization study~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc} demonstrated that the experimental precision depended only weakly on the number of injections $m$, recommending $m = 10$ for processes confidently known to involve 1:1 complexation.
This is in sharp contrast to $\sim$30 injections often recommended by standard protocols in order to visualize a full sigmoidal (S-shaped) curve in the enthalpogram, which would unnecessarily limit precision here by reducing the heat per injection (increasing RSEs to 19\% and 4\%, respectively), as well as increasing the duration of the titration experiment nearly three-fold~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc}.
Using 10 injections, each of volume $v = 10$ $\mu$L, we can compute the approximate syringe concentration $[X]_s$ required from the relevant volumes.
For perfusion-type instruments, assuming instantaneous mixing (appropriate for most reactions~\cite{tellinghuisen:anal-biochem:2007:itc-calibration-nacl}), this is given by,
\begin{equation}
%[X]_s = [M]_0 \, R_m \, V_0 / (m v) \label{equation:syringe-concentration} % old equation
[X]_s = R_m [M]_0 \left[ 1 - \exp\left(-\frac{m v}{V_0}\right)\right]^{-1} \approx R_m [M]_0 \frac{V_0}{m v} \label{equation:syringe-concentration}
\end{equation}
where $V_0$ is the cell active volume ($\sim$1.4 mL for the VP-ITC) and the approximate equality follows if the total titrant injected is small compared to the cell volume ($m v \ll V_0$).
For our experiment, Eq.~\ref{equation:syringe-concentration} suggests we should use a purity-corrected titrant concentration $[X]_s\sim720$~$\mu$M.

Our GE/MicroCal VP-ITC instrument has a syringe assembly that utilizes a worm gear which, after the recommended purge-refill process, will cause a titrant shortfall in the first injection unless a ``down syringe'' command is issued prior to loading the syringe into the sample cell~\cite{tellinghuisen:anal-biochem:2004:first-injection-anomaly}; we therefore executed a 10~$\mu$L ``down syringe'' command immediately after the purge-refill cycle.
Because the instrument can take a substantial period of time to stabilize at the desired experimental temperature after loading the syringe, significant ($>$ 0.1 $\mu$L) diffusive loss can also contribute to a first injection shortfall.
We therefore programmed an initial 1 $\mu$L ``throwaway injection'' to avoid the need to correct for diffusive titrant loss during the first 10 $\mu$L injection.
The contribution from this initial 1 $\mu$L ``throwaway'' injection was excluded from the fitting procedure during analysis.

%\section*{Titrate preparation}

%Bovine carbonic anhydrase (Sigma-Aldrich, cat no.~C2522, 80\% purity) was purchased as a lyophylized powder. 
%Following the assay conditions outlined in Myszka et al.~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}, 2 L of PBS buffer was prepared for overnight dialysis of the protein sample. 
%The protein solution was prepared by dissolving $25\pm0.1$ mg in $4\pm0.1$ mL of PBS buffer. 
%The concentration of the resulting solution was determined prior- and post-dialysis spectrophotometrically on a Hewlett-Packard Agilent UV-Vis spectrophotometer model 8453, using the molar extinction coefficient $\epsilon_{280 \:\mathrm{nm}}$  = 50070 M$^{-1}$ cm$^{-1}$~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}. 
%The sample was then diluted to [M]$_0$ $\sim$ 10 $\mu$M using the purity-corrected post-dialysis concentration. 
%Because the site parameter $n$ absorbs errors in [M]$_0$ during the fitting procedure [Tellinghuisen], high precision is not required for these steps; a less precise spectrophotometer (such as the popular NanoDrop ND-1000) would suffice.

%\section*{Titrant preparation}

The titrand solution, bovine CAII (Sigma-Aldrich, cat no.~C2522, $\sim$29 kDa{\color{red}[JDC: Check MW.]}, Lot No.~071M6261) in PBS buffer, was prepared following the assay conditions outlined by Myszka et al.~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}.  
Briefly, the contents of the glass vial containing $\sim$ 5 mg of lyophilized CAII were resuspended in 750 $\mu$L filtered buffer and dialyzed overnight in 1 L buffer using a Novagen D-Tube Dialyzer MWCO 3.5 kDa (Cat No.~71506-3, Lot D00131446).
The recovered protein was spun for 30 min at 16300 RPM with no visible precipitate observed.
The dialysate was filtered again and used to prepare both titrant and tetrad to minimize buffer mismatch heats during the ITC experiment.

The protein concentration was determined spectrophotometrically via absorbance at 280 nm on a NanoDrop ND-1000.
The NanoDrop (and similar instruments) utilize small sample volumes (3 $\mu$L was used here) and a 1 mm path length such that even high protein concentrations can be quickly read without the need to perform intermediate dilutions to ensure the absorbance falls in the linear range of Beer's law on standard 10 mm path-length spectrometers.
The measured absorbance of 11.77$\pm$0.03 indicated a protein concentration of 235.0$\pm$0.6 $\mu$M (using the molar extension coefficient ($\epsilon_\mathrm{280\:nm} = 50070$ M$^{-1}$ cm$^{-1}$~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}).
{\color{red}[JDC: Should we include the error in $\epsilon$ here?  Or perhaps we should drop the error bars since high precision isn't required?]}
The sample was then diluted to [M]$_0$ $\approx$ 10 $\mu$M using the purity-corrected post-dialysis concentration. 
Note that, because the site parameter $n$ absorbs errors in [M]$_0$ in standard least-squares data analysis~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}, high precision is not required for these steps unless the stoichiometry is unknown.

By contrast, care must be taken to minimize inaccuracies in preparing titrant solutions, because the standard data analysis algorithms treat [X]$_s$ as exactly known.  
Thus, a 1\% error in [X]$_s$ produces 1\% errors in the estimates of $K$ and $\Delta H$~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc,tellinghuisen-chodera:2011:anal-biochem:systematic-itc-errors}.  
Our titrant (CBS, Sigma-Aldrich, lot \#MKBF3323V, 97\% purity, MW 201.2) comes as a powder, from which we aim to prepare a solution of 
purity-corrected\footnote{Recall that 97\% purity denotes 1 g of powder should contain 0.97 g CBS.}  concentration $[X]_s \sim 720$ $\mu$M using the dialysate. 
Uncertainties in the true [X]$_s$ come from at least two sources: the mass of CBS and volume of buffer used in preparing this solution, each of which will be imprecise due to measurement error.  
Further dilution steps will introduce additional error.

@@

To load the VP-ITC syringe, we require $\sim$2.1 mL of our titrant\footnote{A smaller 700 $\mu$L filling tube is also available.}.  
For our chosen [X]$_s$ $\sim$ 720, this requires only 0.3 mg of CBS, but given our analytical balance precision ($\pm$0.1 mg), this would yield 33\% uncertainty in [X]$_s$, and hence the final relative errors in $K$ and $\Delta H$ would be \emph{at least} this large.  
{\color{red}[JDC: Which model balance was used for this?]}
To reduce the mass uncertainty to 1\%, we must weigh out at least 10 mg.  
Since the solubility of CBS in water is only 453 mg/L (which corresponds to a 2 250 $\mu$M solution), we need a volume of at least 22 mL to dissolve 10 mg.
Using a 25 mL Class A volumetric flask or pipette (rated $\pm$0.05 mL) would allow us to attain the desired 1\% precision.  
On the other hand, graduated cylinders and serological pipettes with 25 mL capacity often possess a precision of only $\pm$0.5 ml, which would raise the uncertainty in [X]$_s$ to 2\%.  
Here, we found it convenient to employ multiple liquid transfers with a Gilson P5000 5 mL pipette, which has a stated reliability of $\pm$0.03 mL at full capacity\footnote{For pipettes, the stated systematic error $\delta$ is generally the dominant source of error, and can be considered random error if each pipette is used once.  However, if the same pipette is used to deliver \emph{multiple} aliquots, the random error contribution for the total volume transferred should be estimated from summing the systematic error $\delta$ for each transfer.  For example, if 25 mL are transferred in five transfers of 5 mL using a P5000 ($\delta = 30$ $\mu$L), the random error contribution is $(5)(0.030) = 0.15$ mL.}.

@@

To ensure sufficient $\sim$718.29 $\mu$M titrant to allow for a ligand-into-buffer blank titration and additional experimental replications if needed, we planned to prepare 9 mL of titrant solution.
This is more than necessary, as minimum of 700 $\mu$L/experiment is required for the VP-ITC if the low-volume syringe loading tube is utilized.
Using the Gilson P5000, we then added 4.309(12) mL CBS stock to 4.691(12) mL PBS to obtain a 719(9) $\mu$M CBS titrant (1\% RSE).  
Error propagation was performed automatically by the spreadsheet (Figure \ref{figure:spreadsheet}) using standard techniques (discussed below).
Note that gravimetric solution preparation---in which the mass of both compound and solvent is used to determine the final concentration---could have been used instead, eliminating the need to ensure all solutions are at the appropriate temperature for volumetric glassware or pipettes.
{\color{red}[JDC: Can we say something about volume variation with temperature and how much of a problem this is expected to be?]}
Automated systems for gravimetric solution preparation and concentration error determination are available, though more commonly used in industrial settings.

Alternatively, we could have determined the titrant concentration [X]$_s$ spectrophotometrically using the known extinction coefficient of CBS at 272 nm (reported as $\epsilon_\mathrm{272\:nm} = 1.31(13) \times 10^3$ M$^{-1}$ cm$^{-1}$~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}).
However, since the uncertainty in the absorbance measurement is 1\%, the uncertainty in the extinction coefficient $\epsilon$ (10\%) would dominate the concentration error, resulting in a spectrophotometrically-determined concentration that is uncertain by 10\%. 
Indeed, the concentration we measure in this manner---700(70) $\mu$M {\color{red}[JDC: I have to re-measure the absorbance of the latest titrant solution.]}---is consistent with that determined by mass and volume, but is an order of magnitude more uncertain; had we chosen to use this spectrophotometrically-determined concentration for [X]$_s$, our final uncertainties in $K$ and $\Delta H$ would be at least 10\%.

The titration dataset (Figure~\ref{figure:enthalpogram}) was analyzed using Origin 7.0 (OriginLab Corp.) after subtracting heats obtained from a separate ligand-into-buffer “blank” titration utilizing the same protocol.
%Here, the blank heats were small and uniform and altered the parameter estimates by only about one SE (see Supplementary Material).  
The least-squares (LS) fit of the thermodynamic parameters to the integrated injection heats are shown in the caption.  
Note that the site parameter $n$ absorbs errors in [M]$_0$ and the cell volume V$_0$; if the actual concentration of active macromolecule is of interest, these quantities will require more precision~\cite{tellinghuisen:anal-biochem:2004:volume-errors-in-itc}.
While $\Delta H$ is rather insensitive to errors in the stated cell volume V$_0$ as a result, those errors can have a substantial effect on $K$, so careful calibration of V$_0$ using standard reactions~(e.g.~\cite{tellinghuisen:biophys-chem:2004:barium-crown-ether-itc,tellinghuisen:anal-biochem:2007:itc-calibration-nacl}) is advised if highly accurate quantitation $K$ is necessary~\cite{tellinghuisen:anal-biochem:2004:volume-errors-in-itc}.
{\color{red}[JDC: Joel, is this statement not sufficient, or do you think we really need to include the $V_0$ calibration results?]}

>>>>

\author[mskcc]{Sarah E.~Boyce\fnref{schrodinger}}
%\ead{sarah.boyce@gilead.com}
%\ead[url]{home page}
%\fntext[label2]{label2}
%\cortext[cor1]{cor1}
%\address[qb3]{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720}
%\fntext[gilead]{Current address: Gilead Sciences, Foster City, CA 94404}
\fntext[schrodinger]{Current address: Schr\"odinger, 120 West 45th St, New York, NY 10036}

@@

\author[mskcc]{John D.~Chodera\corref{cor1}}
%\fntext[label2]{label2}
\cortext[cor1]{Corresponding author.}
\ead{john.chodera@choderalab.org}
%\ead[url]{http://choderalab.org}

\address[schrodinger]{Schr\"odinger, 120 West 45th St, New York, NY 10036}
\address[mskcc]{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY}
\address[qb3]{California Institute of Quantitative Biosciences (QB3), University of California, Berkeley, CA 94720}
\address[vanderbilt]{Department of Chemistry, Vanderbilt University, Nashville, TN 37235}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NO ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Isothermal titration calorimetry (ITC) can yield precise ($\pm$3\%) estimates of the thermodynamic parameters describing biomolecular association (affinity, enthalpy, and entropy), making it an indispensable tool for biochemistry and drug discovery.  
Surprisingly, interlaboratory comparisons suggest that errors of $\sim$20\% are common and widely underreported. 
Here, we show how to reduce precision- and accuracy-limiting errors while obtaining good estimates and minimizing material and time consumed by an experiment.
We provide a simple spreadsheet that allows practitioners to identify precision-limiting operations during protocol design, track precision during the experiment, and propagate error to yield realistic final uncertainties.\\
\emph{Keywords: isothermal titration calorimetry (ITC), propagation of error, entropy-enthalpy compensation}
\end{abstract}

@@

{\bf Illustrative application to CAII:CBS.}
For illustration, we consider the target reaction from the ABRF-MIRG'02 survey~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}, the 1:1 association of CBS and bovine CAII, which can be written,
\begin{equation}
M + X \stackrel{K}{\leftrightharpoons} MX \label{equation:binding-reaction} ,
\end{equation}
where $M$ denotes macromolecule and $X$ ligand.
This reaction has a $K_a \sim 10^6$M and $\Delta H \sim -10$ kcal/mol~\cite{myszka:2003:j-biomol-tech:abrf-mirg02,tellinghuisen-chodera:2011:anal-biochem:systematic-itc-errors}.

As both protein and ligand may be precious, there is a desire to minimize material use in protein-ligand studies.
Using concentrations only as large as necessary also minimizes the need for buffer additives such as DMSO to enhance solubility, reducing agents to prevent crosslinking, and detergents to inhibit aggregation. 
These additives pose additional experimental challenges, as calorimetrically-measured heats can be sensitive to even small composition mismatches between cell and syringe solutions. 
Minimizing these effects requires dialysis of the macromolecule by buffer followed by preparation of the ligand in the dialysate.  
If the ligand is already in solution (e.g.~in DMSO stocks), it may not be possible to fully eliminate excipients, leading to potential heat effects due to buffer mismatch even if attempts are made to match compositions.
%~\cite{tellinghuisen-boyce-chodera:2012:anal-biochem:blanks}.

{\bf Experimental design.}
In the ABRF-MIRG'02 survey~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}, participants employed titrand (cell reagent) concentrations [M]$_0$ in the range 7--71 $\mu$M.
We used an ITC protocol design program~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}, which indicated $\sim$3\% relative squared error (RSE) in $K$ and $\sim$1\% for $\Delta H$ was possible with our instrument (a GE/MicroCal VP-ITC) using [M]$_0$ = 10 $\mu$M (consuming $\sim$0.5 mg protein per experiment).
While this gives $c = K [M]_0 \approx 10$, a key ITC parameter~\cite{wiseman:anal-biochem:1989:itc}, in the low range of that recommended by standard protocols, high measurement precision can still be obtained at this $c$ value by titrating to an optimal titrant:titrand ratio $R_m$ given by,
\begin{equation}
R_m = \frac{6.4}{c^{0.2}} +\frac{13}{c} \label{equation:limiting-concentration},
\end{equation}
a heuristic expression\footnote{While use of this expression requires a rough estimate of the reaction $K$ and an $[M]_0$ that will produce observable heats, this is currently unavoidable in the practice of calorimetry. In the worst case, a pilot experiment using minimal material can be used to crudely estimate these quantities and Eq.~\ref{equation:limiting-concentration} used to determine optimal conditions for a second experiment.} obtained from a comprehensive study of precision as a function of $R_m$~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc}.
The suggested $R_m = 5.3$ is significantly greater than the $R_m = 2$ that is widely used in standard protocols for ITC; with decreasing $c$, use of $R_m = 2$ progressively limits the fractional conversion of M to MX and thus limits the precision of estimation for both $K$ and $\Delta H$~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}.
In the present case, use of $R_m = 2$ would cause significant precision loss, almost doubling the achievable RSEs for $K$ and $\Delta H$.

The same optimization study~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc} demonstrated that the experimental precision depended only weakly on the number of injections $m$, recommending $m = 10$ for processes confidently known to involve 1:1 complexation.
This is in sharp contrast to $\sim$30 injections often recommended by standard protocols in order to visualize a full sigmoidal (S-shaped) curve in the enthalpogram, which unnecessarily limits precision by reducing the heat per injection (increasing RSEs to 19\% and 4\%, respectively), as well as increasing the duration of the titration experiment nearly three-fold~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc}.
Using 10 injections, each of volume $v = 10$ $\mu$L, we can compute the approximate syringe concentration $[X]_s$ required from the relevant volumes.
For perfusion-type instruments, assuming instantaneous mixing (appropriate for most reactions~\cite{tellinghuisen:anal-biochem:2007:itc-calibration-nacl}), this is given by,
\begin{equation}
%[X]_s = [M]_0 \, R_m \, V_0 / (m v) \label{equation:syringe-concentration} % old equation
[X]_s = R_m [M]_0 \left[ 1 - \exp\left(-\frac{m v}{V_0}\right)\right]^{-1} \approx R_m [M]_0 \frac{V_0}{m v} \label{equation:syringe-concentration}
\end{equation}
where $V_0$ is the cell active volume ($\sim$1.4 mL for the VP-ITC) and the approximate equality follows if the total titrant injected is small compared to the cell volume ($m v \ll V_0$).
For our experiment, Eq.~\ref{equation:syringe-concentration} suggests we should use a purity-corrected titrant concentration $[X]_s\sim720$~$\mu$M.

{\bf Syringe backlash and the first-injection anomaly.}
Our GE/MicroCal VP-ITC instrument has a syringe assembly that utilizes a worm gear which, after the recommended purge-refill process, will cause a titrant shortfall in the first injection unless a ``down syringe'' command is issued prior to loading the syringe into the sample cell~\cite{tellinghuisen:anal-biochem:2004:first-injection-anomaly}; we therefore executed a 10~$\mu$L ``down syringe'' command immediately after the purge-refill cycle.
Because the instrument can take a substantial period of time to stabilize at the desired experimental temperature after loading the syringe, significant ($>$ 0.1 $\mu$L) diffusive loss can also contribute to a first injection shortfall.
We therefore programmed an initial 1 $\mu$L ``throwaway injection'' to avoid the need to correct for diffusive titrant loss during the first 10 $\mu$L injection.
The contribution from this initial 1 $\mu$L ``throwaway'' injection was excluded from the fitting procedure during analysis.

{\bf Titrand preparation.} 
The titrand solution, bovine CAII (Sigma-Aldrich, cat no.~C2522, $\sim$30 kDa, Lot No.~071M6261) in PBS buffer, was prepared following the assay conditions outlined by Myszka et al.~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}.  
Briefly, the contents of the glass vial containing $\sim$ 5 mg of lyophilized CAII were resuspended in 750 $\mu$L filtered buffer and dialyzed overnight in 1 L buffer using a Novagen D-Tube Dialyzer MWCO 3.5 kDa (Cat No.~71506-3, Lot D00131446).
The recovered protein was spun for 30 min at 16~300 RPM with no visible precipitate observed.
The dialysate was filtered again and used to prepare both titrant and titrand to minimize buffer mismatch heats during the ITC experiment.

The protein concentration was determined spectrophotometrically via absorbance at 280 nm on a NanoDrop ND-1000.
\color{red}
The NanoDrop (and similar instruments) utilize small sample volumes (3 $\mu$L was used here) and a 1 mm path length such that even high protein concentrations can be quickly read without the need to perform intermediate dilutions to ensure the absorbance falls in the linear range of Beer's law on standard 10 mm path-length spectrometers.
The measured absorbance of 11.77$\pm$0.03 indicated a protein concentration of 235.0$\pm$0.6 $\mu$M (using the molar extension coefficient ($\epsilon_\mathrm{280\:nm} = 50070$ M$^{-1}$ cm$^{-1}$~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}).
The sample was then diluted to [M]$_0$ $\approx$ 10 $\mu$M using the purity-corrected post-dialysis concentration. 
Note that, because the site parameter $n$ absorbs errors in [M]$_0$ in standard least-squares data analysis~\cite{tellinghuisen:2012:anal-biochem:designing-itc-experiments}, high precision is not required for these steps unless the stoichiometry is unknown.
\color{black}

{\bf Titrant preparation.}
In contrast to titrand preparation, care must be taken to minimize inaccuracies in preparing titrant solutions, because the standard data analysis algorithms treat [X]$_s$ as exactly known.  
Thus, a 1\% error in [X]$_s$ produces 1\% errors in the estimates of $K$ and $\Delta H$~\cite{tellinghuisen:jpcb:2005:optimizing-experimental-parameters-itc,tellinghuisen-chodera:2011:anal-biochem:systematic-itc-errors}.  
Our titrant (CBS, Sigma-Aldrich, lot \#MKBF3323V, 97\% purity, MW 201.2) comes as a powder, from which we aim to prepare a solution of 
purity-corrected\footnote{Recall that 97\% purity denotes 1 g of powder should contain 0.97 g CBS.}  concentration $[X]_s \sim 720$ $\mu$M using the dialysate. 
Uncertainties in the true [X]$_s$ come from at least two sources: the mass of CBS and volume of buffer used in preparing this solution, each of which will be imprecise due to measurement error.  
Further dilution steps will introduce additional error.

@@

To load the VP-ITC syringe, we require $\sim$2.1 mL of our titrant\footnote{A smaller 700 $\mu$L filling tube is also available.}.  
For our chosen [X]$_s$ $\sim$ 720, this requires only 0.3 mg of CBS, but given the precision of the analytical balance used for this step (Mettler-Toledo AB204, readability $\pm$0.1 mg), this would yield 33\% uncertainty in [X]$_s$, and hence the final relative errors in $K$ and $\Delta H$ would be \emph{at least} this large.  
To reduce the mass uncertainty to 1\%, we must weigh out at least 10 mg.  
Since the solubility of CBS in water is only 453 mg/L (which corresponds to a 2 250 $\mu$M solution), we need a volume of at least 22 mL to dissolve 10 mg.
Using a 25 mL Class A volumetric flask or pipette (rated $\pm$0.05 mL) would allow us to attain the desired 1\% precision.  
On the other hand, graduated cylinders and serological pipettes with 25 mL capacity often possess a precision of only $\pm$0.5 ml, which would raise the uncertainty in [X]$_s$ to 2\%.  
Here, we found it convenient to employ multiple liquid transfers with a Gilson P5000 5 mL pipette, which has a stated reliability of $\pm$0.03 mL at full capacity\footnote{For pipettes, the stated systematic error $\delta$ is generally the dominant source of error, and can be considered random error if each pipette is used once.  However, if the same pipette is used to deliver \emph{multiple} aliquots, the random error contribution for the total volume transferred should be estimated from summing the systematic error $\delta$ for each transfer.  For example, if 25 mL are transferred in five transfers of 5 mL using a P5000 ($\delta = 30$ $\mu$L), the random error contribution is $(5)(0.030) = 0.15$ mL.}.

@@

To ensure sufficient $\sim$718.29 $\mu$M titrant to allow for a ligand-into-buffer blank titration and additional experimental replications if needed, we planned to prepare 9 mL of titrant solution.
This is more than necessary, as minimum of 700 $\mu$L/experiment is required for the VP-ITC if the low-volume syringe loading tube is utilized.
Using the Gilson P5000, we then added 4.309(12) mL CBS stock to 4.691(12) mL PBS to obtain a 719(9) $\mu$M CBS titrant (1\% RSE).  
Error propagation was performed automatically by the spreadsheet (Figure \ref{figure:spreadsheet}) using standard techniques (discussed below).
\color{red}
The use of volumetric glassware in principle requires all solutions and glassware to be equilibrated to the glassware calibration temperature, but due to the low coefficient of thermal expansion of borosilicate glass, this expansion will only introduce an error of 0.0010\%/C~\cite{lembeck:nbsir-74-461:calibration-of-volumetric-glassware}---small enough to be negligible for our purposes.
Note that gravimetric solution preparation (GSP)---in which the mass of both compound and solvent is used to determine the final concentration---could have been used instead, but runs the risk of introducing more significant errors due to the larger coefficient of thermal expansion of liquids (dilute aqueous buffers have a coefficient of thermal expansion near that of water, $\sim$0.025\%/C); however, this still generally amounts to a negligible contribution to error for calorimetry even for changes of several degrees.
Automated systems for gravimetric solution preparation and concentration error determination are available (such as the Mettler-Toledo Quantos), though more commonly used in industrial settings.
\color{black}

Alternatively, we could have determined the titrant concentration [X]$_s$ spectrophotometrically using the known extinction coefficient of CBS at 272 nm (reported as $\epsilon_\mathrm{272\:nm} = 1.31(13) \times 10^3$ M$^{-1}$ cm$^{-1}$~\cite{myszka:2003:j-biomol-tech:abrf-mirg02}).
However, since the uncertainty in the absorbance measurement is 1\%, the uncertainty in the extinction coefficient $\epsilon$ (10\%) would dominate the concentration error, resulting in a spectrophotometrically-determined concentration that is uncertain by $\sim$10\%. 
Indeed, the concentration we measure in this manner---700(70) $\mu$M---is consistent with that determined by mass and volume, but is an order of magnitude more uncertain; had we chosen to use this spectrophotometrically-determined concentration for [X]$_s$, our final uncertainties in $K$ and $\Delta H$ would be at least 10\%.

{\bf Data analysis.}
The titration dataset (Figure~\ref{figure:enthalpogram}) was analyzed using Origin 7.0 (OriginLab Corp.) after subtracting heats obtained from a separate ligand-into-buffer “blank” titration utilizing the same protocol.
%Here, the blank heats were small and uniform and altered the parameter estimates by only about one SE (see Supplementary Material).  
The least-squares (LS) fit of the thermodynamic parameters to the integrated injection heats are shown in the caption.  
Note that the site parameter $n$ absorbs errors in [M]$_0$ and the cell volume V$_0$; if the actual concentration of active macromolecule is of interest, these quantities will require more precision~\cite{tellinghuisen:anal-biochem:2004:volume-errors-in-itc}.
While $\Delta H$ is rather insensitive to errors in the stated cell volume V$_0$ as a result, those errors can have a substantial effect on $K$, so careful calibration of V$_0$ using standard reactions~(e.g.~\cite{tellinghuisen:biophys-chem:2004:barium-crown-ether-itc,tellinghuisen:anal-biochem:2007:itc-calibration-nacl}) is advised if highly accurate quantitation $K$ is necessary~\cite{tellinghuisen:anal-biochem:2004:volume-errors-in-itc}.
Raw and processed datasets are provided with this work as Supplementary Material.

{\bf Error propagation.}
The error reported by the LS fit only represents the error in model fitting assuming the specified concentration for titrant is \emph{exact}---we must now include the uncertainty in the titrant concentration to obtain an estimate of the true error.
The general rule for random error propagation for a quantity $f(x,y,z,\ldots)$ dependent on \emph{independent} measurements $x,y,z,\ldots$ gives squared standard error $s_f$ in $f$ as~\cite{taylor-error-propagation},
\begin{equation}
s_f^2 = \left(\frac{\partial f}{\partial x}\right)^2 s_x^2 + \left(\frac{\partial f}{\partial y}\right)^2 s_y^2 + \left(\frac{\partial f}{\partial z}\right)^2 s_z^2 + \cdots \label{equation:propagation-of-error}
\end{equation}
where $s_x$, $s_y$, and $s_z$ are the standard errors of the corresponding measurements.
This form, based on a Taylor expansion of the function $f$, can be extended to any number of contributing terms (e.g., additional volume contributions if multiple dilutions are used)~\cite{taylor-error-propagation}.  
If $f$ has a simple form, so that it can be written $f(x,y,z,\ldots) = x^i y^j z^k \cdots$, then Eq.~\ref{equation:propagation-of-error} assumes the simple form,
\begin{equation}
\left(\frac{s_f}{f}\right)^2 = i^2 \left(\frac{s_x}{x}\right)^2 + j^2 \left(\frac{s_y}{y}\right)^2 + k^2 \left(\frac{s_y}{z}\right)^2 + \cdots \label{equation:relative-error}
\end{equation}
where $i$, $j$, and $k$ are the powers to which the independent measurements $x$, $y$, and $z$ are raised, and $(s_f / f)$ is the relative error in $f$.
Often, a single term in Eq.~\ref{equation:relative-error} will dominate, and the relative error is essentially identical to this contribution.
For example, if $f = x/y$ with $(s_x/x)$ = 4\% and $(s_y/y)$ = 1\%, then $(s_f/f) = 4.1\% \approx 4\%$.
